Bandit Multiclass Linear Classification: \\ Efficient Algorithms for the Separable Case

### 2019-02-06 arxiv/mistake-lower-bound-for-ignorant-algorithms.tex
elaborating on truncated geom

In rounds $1,2,\dots,\tau-1$, the algorithm predicts the incorrect class $2$
with probability at least $q_0$. Thus,
$$
\Exp\left[\sum_{t=1}^{\tau - 1} \indicator{\widehat y_t\neq y_t}\right] = q_0 \Exp[(\tau - 1)] \; .
$$
In rounds $\tau, \tau+1, \dots, T$, all the examples are the same and are equal
to $(v_\tau, 2)$. Let $s$ be the first time step $t$ such that $t \ge \tau$
and the algorithm makes a correct prediction. If the algorithm makes mistakes
in all rounds $\tau, \tau+1, \dots, T$, we define $s = T+1$.
By definition the algorithm makes mistakes in rounds $\tau, \tau+1, \dots, s-1$.
Therefore,
$$
\Exp\left[\sum_{t=\tau}^T \indicator{\widehat y_t\neq y_t}\right] \ge \Exp[s-\tau].
$$
Since the algorithm is ignorant, conditioned on $\tau$, the random variable $s-\tau$
has a truncated geometric distribution with parameter $p_\tau(2|v_\tau) \le q_0$.
Therefore,
$$
\Exp[s-\tau ~|~ \tau]
\ge
\Exp \left[\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right) ~\middle|~ \tau \right] \; .
$$

@@

Observe that in phase 1, the labels are equal to $1$ and in phase 2 the labels
are equal to $2$. Note that $(x_\tau, y_\tau)=(x_{\tau+1}, y_{\tau+1})= \dots =
(x_T, y_T) = (v_\tau, 2)$. Consider the vectors $u_1, u_2, \dots, u_N$ as
defined in Lemma~\ref{lemma:embed_d_gamma} We claim that $w_1=-u_\tau/2$ and
$w_2=u_\tau/2$ satisfy the conditions of strong linear separability.

@@

>>>>

In rounds $1,2,\dots,\tau-1$, the algorithm predicts the incorrect class $2$
with probability at least $q_0$. Thus,
$$
\Exp\left[\sum_{t=1}^{\tau - 1} \indicator{\widehat y_t\neq y_t}\right] = q_0 \Exp[(\tau - 1)] \; .
$$
In rounds $\tau, \tau+1, \dots, T$, all the examples are the same and are equal
to $(v_\tau, 2)$. Let $s$ be the first time step $t$ such that $t \ge \tau$
and the algorithm makes a correct prediction. If the algorithm makes mistakes
in all rounds $\tau, \tau+1, \dots, T$, we define $s = T+1$.
By definition the algorithm makes mistakes in rounds $\tau, \tau+1, \dots, s-1$.
Therefore,
$$
\Exp\left[\sum_{t=\tau}^T \indicator{\widehat y_t\neq y_t}\right] \ge \Exp[s-\tau].
$$
Since the algorithm is ignorant, conditioned on $\tau$, $s-\tau+1$ has the same distribution as
$\min(X, T-\tau+2)$, where $X$ is drawn from a geometric distribution with parameter
$p_\tau(2|v_\tau)$.

By Lemma~\ref{lem:truncated-geom} below,
\[
  \Exp\left[s-\tau+1 \mid \tau, p_\tau(2|v_\tau) \right] = F(p_\tau(2|v_\tau), T-\tau+2) \geq F(q_0, T-\tau+2).
\]
Therefore,
$$
\Exp[s-\tau ~|~ \tau]
\ge
F(q_0, T-\tau+2) - 1
=
\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right) \; .
$$

%the random variable $s-\tau+1$
%has a truncated geometric distribution with parameter $p_\tau(2|v_\tau) \le q_0$.
%Specifically, conditioned on $\tau$,

@@

Observe that in phase 1, the labels are equal to $1$ and in phase 2 the labels
are equal to $2$. Note that $(x_\tau, y_\tau)=(x_{\tau+1}, y_{\tau+1})= \dots =
(x_T, y_T) = (v_\tau, 2)$. Consider the vectors $u_1, u_2, \dots, u_N$ as
defined in Lemma~\ref{lemma:embed_d_gamma}. We claim that $w_1=-u_\tau/2$ and
$w_2=u_\tau/2$ satisfy the conditions of strong linear separability.

@@

\begin{lemma}
For $p \in (0,1)$ and $N \geq 1$, denote by $F(p, N) = \frac{1}{p}(1 - (1-p)^N)$.
\begin{enumerate}
  \item $F(p, N)$ is monotonically decreasing in $p$.
  \item If $X \sim \Geom(p)$, then $\Exp[\min(X, N)] = F(p, N)$.
\end{enumerate}
\label{lem:truncated-geom}
\end{lemma}
\begin{proof}
We show the two items respectively.
\begin{enumerate}
\item By calculus,
\[ \frac{\partial F(p,n)}{\partial p} = \frac{(1-p)^{N-1} (1 + (N-1)p) - 1}{p^2}. \]
Now, note that for any $p \in (0,1)$,
\[ (1 + (N-1)p) \leq \frac{1}{(1 - p)^{N-1}} \]
We therefore get $\frac{\partial F(p,n)}{\partial p} < 0$ for all $p$, showing the first item.

\item We have the following decomposition on $\Exp[X]$:
\[ \Exp[X] = \Exp[X \one[X \leq N]] + \Prob(X \geq N+1) \Exp[X|X \geq N+1] \]
By the memoryless property of geometric distribution, $\Exp[X|X \geq N+1] = N + \Exp[X] = N + \frac 1 p$.
Therefore,
\[ \Exp[X] = \Exp[X \one[X \leq N]] + \Prob(X \geq N+1) \cdot N + \Prob(X \geq N+1) \cdot \frac 1 p. \]
At the same time, observe that $\Exp[X \one[X \leq N]] + \Prob(X \geq N+1) \cdot N = \Exp[\min(X, N)]$.
Therefore,
\[ \Exp[\min(X, N)] = \frac 1 p \cdot (1 - \Prob(X \geq N+1)) = \frac{1}{p}(1 - (1-p)^N) = F(p, N). \qedhere \]
\end{enumerate}
\end{proof}
